The Talking Heads experiment

### 2015-04-28 chap1/chap1.tex
format relevant papers section

For centuries, philosophers, linguists, psychologists and 
neuroscientists have been groping with the amazing capacities
of the mind. By necessity, they have been doing this through
thought experiments or by observing human behaviour and
brain anatomy. Although this has generated a wealth
of insights,\footnote{Attempts are made to bring together the insights from various
disciplines to establish a true ``cognitive science''. 
\cite{Osherton:1995} contains an introduction into some of the main 
research trends in this very diverse scientific field. 
See also \cite{Luger:1994}. The work reported in the present book can be 
classified as theoretical
cognitive science because I try to formulate and test
the operational adequacy of possible models for the 
origins of language and meaning but do not claim nor
give any evidence that these models are also valid for
human cognition, just as the study of aerodynamics and 
aircraft design may help to understand how birds can 
fly but is more generic than its biological 
implementations.}
everyone involved in this research must surely 
agree that we are still lacking adequate models, 
particularly for higher order cognitive functions like language
processing, and definitely for understanding how such 
functions may have arisen. 
To discover or test such models, it therefore
remains useful to do experiments with artificial systems.
We can build robots that receive sensory inputs
through a camera or other sensors, give them computational 
power and memory, and empower them 
for action in the world by adding actuators. Given such a 
set-up, we can precisely examine the operational adequacy 
of a hypothesis. For example, if someone proposes
a process for segmenting images, we can test this process 
by capturing streams of images through a camera and 
see whether the process is indeed capable of performing 
segmentation. When someone proposes a process 
for parsing sentences, we 
can implement this process and confront it with a series of 
example sentences to measure success and failure. 
How else can we test the operational adequacy of a proposed
cognitive model, given the enormous complexity involved? 
Of course, building an artificial system in no way proves
that the principles that were used to construct it are valid
for natural systems. But it is an enormously valuable 
source of insight for approaching the extraordinarily 
complex phenomena observed in human cognition.

@@

The Talking Heads experiment follows this research strategy. \is{Talking Heads Experiment}
It features an enormously challenging experimental
infrastructure to explore how a cognitive
system, like the one underlying human language, might
be able to bootstrap itself into interaction 
with other cognitive systems and driven by increasing
challenges from the environment. The experiment
involves a set of robotic ``Talking Heads'' engaged in language
games, with each other or with human interlocutors, 
about real world scenes they perceive through their sensors
(see Figures \ref{f:plate1} and \ref{f:plate2}).
The robots use vision as major sensory source. They are located in
different places in the world and connected through the Internet. 
Two robotic agents can only engage in an interaction when they 
are instantiated in robot bodies in a shared physical environment. 
After an exchange, an agent can teleport himself to another body in 
another location and engage in interactions there.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{./chap1/figs/head.pdf}}
\caption{A single ``Talking Head''. There is a camera oriented towards a
white board. The bottom screen shows what the camera observes. The top right screen shows the result of processing.
A loudspeaker reproduces the utterances of the agent. }
\label{f:plate2}
\end{figure}

@@

The second radical idea is that 
meaning is built up slowly by each individual in a cumulative
growth process. Meaning is not innate, as rationalists in the 
footprints of Plato have been arguing for centuries, nor 
learned through stepwise
induction from examples and counterexamples, as empiricists
have been saying. Meaning is at first very concrete and
strongly situated in the environment and bodily experiences.
I will take the suggestions made
by Wittgenstein \is{Wittgenstein} one step further, namely that meaning (and language) is constructed
and practised as part of
language games.\footnote{\cite{Wittgenstein:1953} emphasises the relativity 
of concepts and the role of language and hence
meaning in social interactions.}
I will introduce a selectionist approach to
the acquisition of meaning, introducing models to 
show that conceptual distinctions can `grow' in the brain
like the leaves and branches on a tree and be pruned to fit the
demands and characteristics of the environment in which
an agent finds itself. Even though non-verbal 
activities, like predicting the future based on a model 
or deciding what to do in specific circumstances, stimulates
the growth of distinctions, language use is probably one 
of the greatest stimulators of conceptual growth. It provides
feedback about which distinctions were successful in linguistic
communication and thus whether or not they should be preserved. 
Thus language and cognition co-evolve. Each one pushes the other 
up towards more complexity and they become tightly 
co-ordinated with neither a central co-ordinator nor prior innate design.

@@

The Talking Heads experiment is the culmination of one of
the most exciting scientific and engineering projects I have
ever been involved in. It has required the
creative efforts of a dozen excellent researchers 
over many years. The story started in 
1985. Instead of continuing to design and program intelligence
explicitly based on formalising human cognitive
capacities, as most of my colleagues
in artificial intelligence research 
labs were doing,\footnote{
Some recent overviews of this `classical' approach to 
artificial intelligence can be found in: \cite{Nilsson:1998} and 
\cite{Russell:1998}.}
I started to focus on the 
question of how intelligence might
originate and evolve in physical agents as they 
interact autonomously with their environment or with 
other humans, and I encouraged my students
at the Artificial Intelligence Laboratory of the University 
of Brussels (VUB) to experiment in the same direction. 
We initially developed a bottom-up, 
behaviour-oriented approach to sensori-motor intelligence, 
which was also being explored around the same time by 
Rodney Brooks at the MIT Artificial Intelligence 
Laboratory.
The behaviour-oriented, bottom-up approach \is{behaviour-based robots} was a 
counter reaction to the symbolic, top-down approach of 
earlier AI research. See \cite{Steels:1995} and \cite{Arkin:1998}. 
We built robots of various sizes and shapes, using simple
electronic circuits, Lego bricks, small motors, rechargeable
batteries, self-made 
sensors, single board computers, and everything else that 
appeared useful (\figref{f:plate3}).

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap1/figs/fish.pdf}}
\caption{``Artificial fish'' built by Miles Pebody to explore influence of robot bodies on behaviour. The fish could swim around by wagging 
its tail and avoid obstacles based on infrared sensing.}
\label{f:plate4}
\end{figure}

@@

Together with 
other researchers interested in the then-arising field of `artificial life', I had 
already been simulating path formation in 
ant societies and other biological phenomena exhibiting 
an emergence of complexity. It dawned on me that the
importance of these mechanisms for bootstrapping intelligence and
language might be much greater than 
thought so far.\footnote{The following reference provides a general survey of similar work in 
the area of lexicon formation: \cite{Steels:97b}. 
A representative sample of work on syntax is \cite{Briscoe:1999}.}

>>>>

For centuries, philosophers, linguists, psychologists and 
neuroscientists have been groping with the amazing capacities
of the mind. By necessity, they have been doing this through
thought experiments or by observing human behaviour and
brain anatomy. Although this has generated a wealth
of insights,\footnote{Attempts are made to bring together the insights from various
disciplines to establish a true \dblenquote{cognitive science}. 
\cite{Osherton:1995} contains an introduction into some of the main 
research trends in this very diverse scientific field. 
See also \cite{Luger:1994}. The work reported in the present book can be 
classified as theoretical
cognitive science because I try to formulate and test
the operational adequacy of possible models for the 
origins of language and meaning but do not claim nor
give any evidence that these models are also valid for
human cognition, just as the study of aerodynamics and 
aircraft design may help to understand how birds can 
fly but is more generic than its biological 
implementations.}
everyone involved in this research must surely 
agree that we are still lacking adequate models, 
particularly for higher order cognitive functions like language
processing, and definitely for understanding how such 
functions may have arisen. 
To discover or test such models, it therefore
remains useful to do experiments with artificial systems.
We can build robots that receive sensory inputs
through a camera or other sensors, give them computational 
power and memory, and empower them 
for action in the world by adding actuators. Given such a 
set-up, we can precisely examine the operational adequacy 
of a hypothesis. For example, if someone proposes
a process for segmenting images, we can test this process 
by capturing streams of images through a camera and 
see whether the process is indeed capable of performing 
segmentation. When someone proposes a process 
for parsing sentences, we 
can implement this process and confront it with a series of 
example sentences to measure success and failure. 
How else can we test the operational adequacy of a proposed
cognitive model, given the enormous complexity involved? 
Of course, building an artificial system in no way proves
that the principles that were used to construct it are valid
for natural systems. But it is an enormously valuable 
source of insight for approaching the extraordinarily 
complex phenomena observed in human cognition.

@@

The Talking Heads experiment follows this research strategy.  
It features an enormously challenging experimental
infrastructure to explore how a cognitive
system, like the one underlying human language, might
be able to bootstrap itself into interaction 
with other cognitive systems and driven by increasing
challenges from the environment. The experiment
involves a set of robotic \dblenquote{Talking Heads} engaged in language
games, with each other or with human interlocutors, 
about real world scenes they perceive through their sensors
(see Figures \ref{f:plate1} and \ref{f:plate2}).
The robots use vision as major sensory source. They are located in
different places in the world and connected through the Internet. 
Two robotic agents can only engage in an interaction when they 
are instantiated in robot bodies in a shared physical environment. 
After an exchange, an agent can teleport himself to another body in 
another location and engage in interactions there.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{./chap1/figs/head.pdf}}
\caption{A single \dblenquote{Talking Head}. There is a camera oriented towards a
white board. The bottom screen shows what the camera observes. The top right screen shows the result of processing.
A loudspeaker reproduces the utterances of the agent. }
\label{f:plate2}
\end{figure}

@@

The second radical idea is that 
meaning is built up slowly by each individual in a cumulative
growth process. Meaning is not innate, as rationalists in the 
footprints of Plato have been arguing for centuries, nor 
learned through stepwise
induction from examples and counterexamples, as empiricists
have been saying. Meaning is at first very concrete and
strongly situated in the environment and bodily experiences.
I will take the suggestions made
by Wittgenstein \is{Wittgenstein} one step further, namely that meaning (and language) is constructed
and practised as part of
language games.\footnote{\cite{Wittgenstein:1953} emphasises the relativity 
of concepts and the role of language and hence
meaning in social interactions.}
I will introduce a selectionist approach to
the acquisition of meaning, introducing models to 
show that conceptual distinctions can \smplenquote{grow} in the brain
like the leaves and branches on a tree and be pruned to fit the
demands and characteristics of the environment in which
an agent finds itself. Even though non-verbal 
activities, like predicting the future based on a model 
or deciding what to do in specific circumstances, stimulates
the growth of distinctions, language use is probably one 
of the greatest stimulators of conceptual growth. It provides
feedback about which distinctions were successful in linguistic
communication and thus whether or not they should be preserved. 
Thus language and cognition co-evolve. Each one pushes the other 
up towards more complexity and they become tightly 
co-ordinated with neither a central co-ordinator nor prior innate design.

@@

The Talking Heads experiment is the culmination of one of
the most exciting scientific and engineering projects I have
ever been involved in. It has required the
creative efforts of a dozen excellent researchers 
over many years. The story started in 
1985. Instead of continuing to design and program intelligence
explicitly based on formalising human cognitive
capacities, as most of my colleagues
in artificial intelligence research 
labs were doing,\footnote{
Some recent overviews of this \smplenquote{classical} approach to 
artificial intelligence can be found in: \cite{Nilsson:1998} and 
\cite{Russell:1998}.}
I started to focus on the 
question of how intelligence might
originate and evolve in physical agents as they 
interact autonomously with their environment or with 
other humans, and I encouraged my students
at the Artificial Intelligence Laboratory of the University 
of Brussels (VUB) to experiment in the same direction. 
We initially developed a bottom-up, 
behaviour-oriented approach to sensori-motor intelligence, 
which was also being explored around the same time by 
Rodney Brooks at the MIT Artificial Intelligence 
Laboratory.
The behaviour-oriented, bottom-up approach \is{behaviour-based robots} was a 
counter reaction to the symbolic, top-down approach of 
earlier AI research. See \cite{Steels:1995} and \cite{Arkin:1998}. 
We built robots of various sizes and shapes, using simple
electronic circuits, Lego bricks, small motors, rechargeable
batteries, self-made 
sensors, single board computers, and everything else that 
appeared useful (\figref{f:plate3}).

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap1/figs/fish.pdf}}
\caption{\dblenquote{Artificial fish} built by Miles Pebody to explore influence of robot bodies on behaviour. The fish could swim around by wagging 
its tail and avoid obstacles based on infrared sensing.}
\label{f:plate4}
\end{figure}

@@

Together with 
other researchers interested in the then-arising field of \smplenquote{artificial life}, I had 
already been simulating path formation in 
ant societies and other biological phenomena exhibiting 
an emergence of complexity. It dawned on me that the
importance of these mechanisms for bootstrapping intelligence and
language might be much greater than 
thought so far.\footnote{The following reference provides a general survey of similar work in 
the area of lexicon formation: \cite{Steels:97b}. 
A representative sample of work on syntax is \cite{Briscoe:1999}.}
